---
title: "Introduction to midnight: Core Modeling and Fast Solvers"
description: >
  Discover the core functionalities of the midnight package, featuring the mid_reg() tidymodels interface and high-performance fastLmMatrix solvers.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

pkgs <- c("midr", "colormap", "DT", "ggplot2", "parsnip", "patchwork")
install.packages(pkgs[!pkgs %in% list.files(.libPaths())])
```

The **midnight** package is designed to provide robust modeling capabilities and highly optimized solvers, working seamlessly alongside the **midr** package.
The core philosophy of **midnight** revolves around its two main functionalities:

1. `mid_reg()`: A seamless **tidymodels** (**parsnip**) interface for intuitive model specification.

2. `fastLmMatrix()`: High-performance multivariate least squares solvers designed for speed and efficiency.

In addition to these core computational features, the package offers advanced visualization tools and integrated color themes.

## The Tidymodels Interface

**midnight** fully integrates with the **tidymodels** ecosystem.
The `mid_reg()` function allows you to specify models using the familiar **parsnip** syntax, enabling pipeline-based workflows for model fitting and hyperparameter tuning.

```{r mid_reg, warning=FALSE}
library(midr)
library(midnight)
library(parsnip)
library(ggplot2)
library(patchwork)

# fitting a model with interaction using mid_reg()
mid <- mid_reg(
  params_main = 50, # k
  penalty = 5 # lambda
) |>
  fit(Ozone ~ (Wind + Solar.R) ^ 2, na.omit(airquality))

mid
```

## High-Performance Least Square Solvers

For situations requiring fitting a surrogate model for multivariate responses, **midnight** provides highly optimized OLS solvers.
You can compare standard **RcppEigen** implementations with the matrix-based solvers provided by **midnight**.

```{r}
# standard RcppEigen solver
fit1 <- RcppEigen::fastLm(
  Ozone ~ Wind + Solar.R, na.omit(airquality)
)
print(as.matrix(fit1$coefficients))
fit2 <- RcppEigen::fastLm(
  Temp ~ Wind + Solar.R, na.omit(airquality)
)
# fastLmMatrix
print(as.matrix(fit2$coefficients))
fit3 <- fastLmMatrix(
  cbind(Ozone, Temp) ~ Wind + Solar.R, na.omit(airquality)
)
print(fit3$coefficients)
```

Furthermore, you can globally override the default Least Squares Solvers used in `mid_reg()` (and `interpret()`) to leverage this speed boost across your entire modeling pipeline by using `set_fastLmMatrix()`.

```{r}
set_fastLmMatrix() # override Least Squares Solvers of midr via options

mids <- mid_reg(
  params_main = 50, # k
  penalty = 5 # lambda
) |>
  fit(cbind(Ozone, Temp) ~ (Wind + Solar.R) ^ 2, airquality)

mids
```

## New Visualization Styles

To understand the output of `mid_reg()`, **midnight** provides S3 methods for MID models like `ggmid.midimp()` and `persp.mid()`.
These methods help translate complex model outputs into interpretable insights.

### Variable Importance

Using `ggmid(type = 'beeswarm')`, as well as `'violinplot'` or `'sinaplot'`, you can effortlessly plot feature importance distributions to understand the impact of individual variables and their interactions.

```{r midimp}
imp <- mid.importance(mid$fit, data = airquality)
p1 <- ggmid(imp, type = "beeswarm", theme = "Hokusai3") +
  theme(legend.position = "bottom")
p2 <- ggmid(imp, type = "violin", theme = "moon") +
  theme(legend.position = "bottom")
p1 + p2
```

### Joint Effect Surface

The S3 method of `persp()` for "mid" creates surface plots, making it easy to examine the joint effect of two predictors.

```{r persp}
par.midr(mar = c(0, 0, 0, 0))
persp(mid$fit, "Wind:Solar.R", theta = 45, phi = 30)
```

## Additional Color Themes

To ensure the visualizations are both accurate and aesthetically pleasing, **midnight** adds a wide array of color palettes from **colormap**, *DALEX* and **MetBrewer** as well as three original themes.

```{r color.theme.info, echo = FALSE}
.df <- color.theme.info()
.df <- .df[.df$source %in% c("midnight", "DALEX", "colormap", "MetBrewer"), ]
DT::datatable(
  .df,
  class = 'cell-border stripe hover',
  rownames = FALSE,
  options = list(
    pageLength = 10,
    lengthMenu = c(10, 20, 50, 100),
    dom = 'Btip'
  )
)
```

### Qualitative Themes

```{r qualitative, echo = FALSE, fig.height=2.2, fig.width=6}
par.midr(mfrow = c(2, 2))
df <- .df[.df$type == "qualitative", ]
for (i in seq_len(nrow(df))) {
  ct <- color.theme(df[i, 1L], source = df[i, 2L])
  text <- sprintf("%s/%s", df[i, 2L], df[i, 1L])
  plot(ct, text = text)
}
```

### Sequential Themes

```{r sequential, echo = FALSE, fig.height=2.2, fig.width=6}
par.midr(mfrow = c(2, 2))
df <- .df[.df$type == "sequential", ]
for (i in seq_len(nrow(df))) {
  ct <- color.theme(df[i, 1L], source = df[i, 2L])
  text <- sprintf("%s/%s", df[i, 2L], df[i, 1L])
  plot(ct, text = text)
}
```

### Diverging Themes

```{r diverging, echo = FALSE, fig.height=2.2, fig.width=6}
par.midr(mfrow = c(2, 2))
df <- .df[.df$type == "diverging", ]
for (i in seq_len(nrow(df))) {
  ct <- color.theme(df[i, 1L], source = df[i, 2L])
  text <- sprintf("%s/%s", df[i, 2L], df[i, 1L])
  plot(ct, text = text)
}
```

## Summary

The **midnight** package combines the computational speed of `fastLmMatrix()` with the modern API of **tidymodels** via `mid_reg()`.
When paired with its powerful visualization and color theme systems, it provides a comprehensive toolkit for building, scaling, and interpreting sophisticated models.
